---
title: "Assignment C-programming "
author: "david redmond ID 14207377"
date: "30 July 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#  Objectives
The code implement 3 numerical root finding algorithms ,  Bisection method ,NewtonRaphson’s method and Secant method

## Methods of operation:
### Bisection
This method is a simple root-finding algorithm for a function f being a continuous function, for which one knows an interval [a, b] such that f(a) and f(b) have opposite signs (a bracket). This is implemented simply by (f(a) * f(b) < 0 ) and one has divided by two the size of the interval. 
Each iteration performs these steps:
1. Calculate c, the midpoint of the interval, c = a + b/2. 
1. Calculate the function value at the midpoint, f(c). 
1. If convergence is satisfactory (that is, a - c is sufficiently small, or f(c) is sufficiently small), return c and stop iterating. 
1. Examine the sign of f(c) and replace either (a, f(a)) or (b, f(b)) with (c, f(c)) so that there is a zero crossing within the new interval. 


### NewtonRaphson's
This method assumes the function f to have a continuous derivative. Newton's method may not converge if started too far away from a root. However, when it does converge, it is faster than the bisection method, and is usually quadratic. Newton's method is also important because it readily generalizes to higher-dimensional problems.
  $$ x(n+1) = x(n) - \frac{f'(x(n))}{f"(x(n))}$$

### Secant method, replacing the derivative in Newton's method with a finite difference, we get the secant method. 

$$ x[n] = x[n-1] - f(x[n-1]) * \frac{(x[n-1] - x[n-2])}{(f(x[n-1]-f(x[n-2]))}$$



##Code structure :
Each algorithm is coded as per recurssion formula, and each functions will have a specific exit to the iteration based on the supplied tolerance value, or maximum number of iterations allowed, or if the exit condition is satisifed. The function also includes a check for convergence for Secant and Newton. The library incl, contains 4 functions : $$1st$$ and $$2nd$$ derivatives log-likelihood for Cauchy PDF, and a min and max value function which is used for the initial guess in the algorithms

There is an exception handler __check_valid_inputs__ that looks for valid inputs for number of iterations, and valid error tolerence. 

Included is a sample test vector 
```{r }
xx <- c( 12.262307 , 10.281078 , 10.287090 , 12.734039 ,
         11.731881 , 8.861998 , 12.246509 , 11.244818 ,
         9.696278 , 11.557572 , 11.112531 , 10.550190 ,
         9.018438 , 10.704774 , 9.515617 , 10.003247 ,
         10.278352 , 9.709630 , 10.963905 , 17.314814)
```


## Benchmarking
Given the following identical conditions each algorithm was benchmarked using the rbenchmark library 
```{r eval= FALSE}

itr=200
tol = 1e-7
inita = 0.0
initb = 1.0
bisectC(xx, itr, tol, inita, initb)
inita = bisectC(xx, itr, tol, inita, initb)
secantC(xx, itr, tol, inita)
NewRapC(xx, itr, tol, inita)
#
benchmark(bisectC(xx, itr, tol, inita, initb),
          secantC(xx, itr, tol,inita),
          NewRapC(xx, itr, tol,inita),
          order = "relative")[,1:4]
```
 Notice that the output of Bisect is used as the initial guess for Secant and Newton. This is because these 2 algorithms are not gauranteed to converge.
 
 